SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2014-05-03 paper/paper.tex
paper: with updated data.

\begin{document}

@@

\todo{need to test with arbitrary for ints turned off.  (Use newtype)}

\begin{figure}[ht]
\begin{code}
type I   = [Int16]
data T   = T (Word8) I I I I

toList :: T -> [[Int16]]
toList (T w i0 i1 i2 i3) =
  [fromIntegral w] : (map . map) fromIntegral rst
  where
  rst = [i0, i1, i2, i3]

@@

\todo{redo analysis, check text}
In this section, we motivate in more detail the challenges in shrinking
counterexamples (we focus on shrinking and only touch on generalization here).
Consider the example in Figure~\ref{fig:initial}.\footnote{All examples and
  algorithms in this paper are presented in Haskell 2010~\cite{haskell2010} plus
  the language extensions of existential type quantification and scoped type
  variables.  We use Haskell \ttp{Prelude} functions, \ttp{mapMaybe} from
  \ttp{Data.Maybe}, and \ttp{liftM} and \ttp{replicateM} from
  \ttp{Control.Monad}.}  Data type \ttp{T} is a product type containing four
lists of wrapped \ttp{Int16}s and a single wrapped \ttp{Word8}.  For our
purposes, it is not important what \ttp{T} models, but for concreteness, suppose
it models the values in four linked lists in an operating system, together with
an 8-bit status register.

Now suppose we are modeling some program that serializes values of type \ttp{T}.
The input to the program satisfies the invariant \ttp{pre}, that the sum of
values in each list of \ttp{Int16}s is less than or equal to 256 (and by
definition, the \ttp{Word8} value is less than or equal to 256).  Assuming this,
we want to show \ttp{post} holds, that the sum of all the values from \ttp{T} is
less than $5 * 256$, where five is the number of fields in \ttp{T}.  At first
glance, the property seems reasonable.  But we have forgotten about underflow;
for example, the value
%
\begin{code}
T 10 [-20000] [-20000] [] []
\end{code}
%
\noindent
satisfies \ttp{pre} but fails \ttp{post} (the \ttp{==>} operator in the figure
is implication from the QuickCheck library).

Despite the simplicity of the example, a typical counterexample returned by
QuickCheck can be large.  With standard settings and no shrinking, the median
counterexample discovered contains around 70 \ttp{Int16} values, and
counterexamples can contain over 100 values.  \todo{recheck} Thus, it pays to
define \ttp{shrink}!

We might first naively try to shrink counterexamples for a data type like
\ttp{T} by taking the cross-product of shrunk values over the arguments to the
constructor \ttp{T}.  This can be expressed using Haskell's list-comprehension
notation:
%
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- shrink w,  b <- shrink i0
                , c <- shrink i1, d <- shrink i2
                , e <- shrink i3 ]
\end{code}
%
\noindent
While the above definition appears reasonable, there are two problems with it.
First, the result of (\ttp{shrink t}) is null if any list contained in \ttp{t}
is null, in which case \ttp{t} will not be shrunk.  More troublesome is that with
QuickCheck's default \ttp{Arbitrary} instances, the length of potential
counterexamples returned by \ttp{shrink} can easily exceed $10^{10}$, which is
an intractable number of tests for QuickCheck to analyze.  The reason for the
blowup is that shrinking a list \ttp{[a]} produces a list \ttp{[[a]]}, the
length of which is significantly longer than the original list.  Then, we take
the cross-product of the generated lists.  For the example above, with some
counterexamples, the shrinking stage may appear to execute for hours, consuming
ever more memory, without returning a result.

A programmer might try to control the complexity by truncating lists using the
(\ttp{take n}) function that returns the first $n$ elements of a list.  The
trade-off is quicker shrinking with a lower probability of finding a smaller
counterexample.  For example, we might redefine shrink as follows:
%
\begin{samepage}
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- tk w
                , b <- tk i0, c <- tk i1
                , d <- tk i2, e <- tk i3 ]
  where tk x = take 10 (shrink x)
\end{code}
\end{samepage}
%
\noindent
Call this version ``truncated shrink''.  Truncation controls the blowup of the
input-space; the maximum number of possible new values is $10^5$ in this case.
While truncation controls the blowup, the downside is that potentially smaller
counterexamples may be omitted.

A more clever programmer that understands the semantics of QuickCheck's
\ttp{shrink} implementation\footnote{The following approach was suggested to
  the author by John Hughes.} defines a \ttp{shrink} instance as follows:
%
\begin{code}
shrink (T w i0 i1 i2 i3) = map go xs
  where xs = shrink (w, i0, i1, i2, i3)
        go (w', i0', i1', i2', i3')
           = T w' i0' i1' i2' i3'
\end{code}
%
\noindent
This ``tuple shrink'' definition does not suffer the same shortcomings: it
shrinks a value even if it contains an empty list, and the combinatorial blowup
of shrink candidates is avoided, since a pair \ttp{(a,b)} is shrunk by
attempting to shrink \ttp{a} while holding \ttp{b} constant, then attempting to
shrink \ttp{b} while holding \ttp{a} constant (and is generalized to larger
tuples).  Thus, each argument to \ttp{T} is independently shrunk.

Still, using this definition, from some initial counterexamples, shrinking may
take over one minute and result in a final counterexample containing just under
100 \ttp{Int16} values.\footnote{All results reported in the paper are with a
  GHC-7.6.3-compiled program, using \ttp{-O2}, on a 4-core 2.7GHz Intel~i7
  processor.}

@@

\todo{QC without shrinking Ints}

\begin{figure}[ht]
\includegraphics[scale=0.6]{../regression/PaperExample1/out/data}
\includegraphics[scale=0.6]{../regression/PaperExample1/out/time-big}
\includegraphics[scale=0.6]{../regression/PaperExample1/out/time-small}
  \caption{Results for 2000 tests of the program in Figure~\ref{fig:initial}.}
  \label{fig:graphs}
\end{figure}

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline
           & QC none    & QC trunc   & QC tuple   & SmartCheck\\
\hline \hline
Mean       & 0.08s, 69  & 0.53s, 35  & 2.24s, 12  & 0.10s, 7\\
\hline
Median     & 0.07s, 69  & 0.17s, 33  & 1.81s, 13  & 0.10s, 6\\
%% \hline
%% MAD        & 0.02s, 13  & 0.09s, 8   & 0.62s, 3   & 0.02s, 2\\
\hline
95\%       & 0.14s, 100 & 1.76s, 65  & 4.06s, 19  & 0.18s, 12\\
\hline
    \end{tabular}
  \end{center}
  \caption{Summarizing data for the graphs in Figure~\ref{fig:graphs}. Entries
    contain execution time (in seconds) and counterexample sizes (counting
    \ttp{N} constructors).}
  \label{table:results}
\end{table}

To motivate the benefits of SmartCheck reduction algorithm, consider
Table~\ref{table:results} and the corresponding graphs in
Figure~\ref{fig:graphs}, giving the results for using QuickCheck and SmartCheck
on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck, since as
noted, it is not even feasible for the example with default instances).  We
graph the final counterexample size and execution time.  Out of 2000 runs, we
show the number of runs resulting in a shrunk counterexample of a given size (by
counting \ttp{N} constructors), and we show the number of runs executing within
a given time limit (in seconds).  Note that we provide two execution-time
graphs, since time scales differ dramatically (we use a logrithmic scale for the
$y$ axis in one graph and a linear scale in the other).  In the table, we show
the mean, median, median, and the results at the 95th percentile values.  (We
give standard deviations, but not that the plots are not necessarily Gaussian.)

The experiments show three approaches to shrinking with QuickCheck: no shrinking
(QC none), to provide a baseline, truncated shrinking (QC trunc), as described
above, in which each list is truncated to 10 elements, and tuple shrinking (QC
tuple), also described above.  SmartCheck produces smaller counterexamples than
the other approaches, with a very small tail.

@@

\multirow{4}{*}{Reverse}
  & Mean       & 0.08s, 69  & 0.53s, 35  \\
  \cline{2-4}
  & Median     & 0.07s, 69  & 0.17s, 33  \\
  \cline{2-4}
  & Std. dev.  &            &            \\
  \cline{2-4}
  & 95\%       & 0.14s, 100 & 1.76s, 65  \\
  \hline

\multirow{4}{*}{Div0}
  & Mean       & 0.08s, 69  & 0.53s, 35  \\
  \cline{2-4}
  & Median     & 0.07s, 69  & 0.17s, 33  \\
  \cline{2-4}
  & Std. dev.  &            &            \\
  \cline{2-4}
  & 95\%       & 0.14s, 100 & 1.76s, 65  \\
  \hline

\multirow{4}{*}{Heap}
  & Mean       & 0.08s, 69  & 0.53s, 35  \\
  \cline{2-4}
  & Median     & 0.07s, 69  & 0.17s, 33  \\
  \cline{2-4}
  & Std. dev.  &            &            \\
  \cline{2-4}
  & 95\%       & 0.14s, 100 & 1.76s, 65  \\
  \hline
    \end{tabular}

@@

@@

\section*{Acknowledgments}
I thank Andy Gill, Joe Hurd, John Launchbury, Simon Winwood, and especially John
Hughes for reading an earlier draft of this paper.  Their deep insights and bug
catches dramatically improved both the form and content.  This paper was and
rejected from ICFP'13; the anonymous reviewers' suggestions improved the
presentation additional benchmarking.

>>>>

\newcommand{\sci}[2]{\ensuremath{#1 \times 10^{#2}}}
\begin{document}

@@

\begin{figure}[ht]
\begin{code}
type I   = [Int16]
data T   = T I I I I I

toList :: T -> [[Int16]]
toList (T i0 i1 i2 i3 i4) =
  (map . map) fromIntegral [i0, i1, i2, i3, i4]

@@

In this section, we motivate in more detail the challenges in shrinking
counterexamples by comparing manual approaches using QuickCheck to SmartCheck.
(We focus on shrinking rather than generalization here since counterexample
generalization is unique to SmartCheck.)  We will show how a small data type and
simple property can result in large counterexamples without any shrinking.  Then
we show the difficulty in designing an efficient shrink implementation manually,
which is required when using QuickCheck; we will walk through a couple of poor
designs before arriving at a ``canonical'' manual solution.  However, even the
canonical solution produces significantly larger counterexamples and takes
significantly longer to produce them than SmartCheck, even though SmartCheck's
shrinking strategy is completely automatic and generic.

Consider the example in Figure~\ref{fig:initial}.\footnote{All examples and
  algorithms in this paper are presented in Haskell 2010~\cite{haskell2010} plus
  the language extensions of existential type quantification and scoped type
  variables.}  Data type \ttp{T} is a product type containing five lists of
signed 8-bit integers.  For our purposes, it is not important what \ttp{T}
models, but for concreteness, suppose it models the values in five linked lists.

Now suppose we are modeling some program that serializes values of type \ttp{T}.
The input to the program satisfies the invariant \ttp{pre}, that the sum of
values in each list of \ttp{Int16}s is less than or equal to 256.  Assuming this,
we want to show \ttp{post} holds, that the sum of all the values from \ttp{T} is
less than $5 * 256$, where five is the number of fields in \ttp{T}.  At first
glance, the property seems reasonable.  But we have forgotten about underflow;
for example, the value
%
\begin{code}
T [-20000] [-20000] [] [] []
\end{code}
%
\noindent
satisfies \ttp{pre} but fails \ttp{post} (the \ttp{==>} operator in the figure
is implication from the QuickCheck library).

Despite the simplicity of the example, a typical counterexample returned by
QuickCheck can be large.  With standard settings and no shrinking, the median
counterexample discovered contains just under 70 \ttp{Int16} values, and
counterexamples can contain over 100 values.  Thus, it pays to define
\ttp{shrink}!

We might first naively try to shrink counterexamples for a data type like
\ttp{T} by taking the cross-product of shrunk values over the arguments to the
constructor \ttp{T}.  This can be expressed using Haskell's list-comprehension
notation:
%
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- shrink w,  b <- shrink i0
                , c <- shrink i1, d <- shrink i2
                , e <- shrink i3 ]
\end{code}
%
\noindent
While the above definition appears reasonable on first glance, there are two
problems with it.  First, the result of (\ttp{shrink t}) is null if any list
contained in \ttp{t} is null, in which case \ttp{t} will not be shrunk.  More
troublesome is that with QuickCheck's default \ttp{Arbitrary} instances, the
length of potential counterexamples returned by \ttp{shrink} can easily exceed
$10^{10}$, which is an intractable number of tests for QuickCheck to analyze.
The reason for the blowup is that shrinking a list \ttp{[a]} produces a list
\ttp{[[a]]}, a list of lists; each list is a new shrinking of the original list.
Then, we take the cross-product of the generated lists.  For the example above,
with some counterexamples, the shrinking stage may appear to execute for hours,
consuming ever more memory, without returning a result.

The first way we will control the complexity is focusing on structurally
shrinking the data type rather than shrinking the base types.  We prevent
QuickCheck from using its default shrink instance for \ttp{Int16} by defining a
newtype:
%
\begin{code}
newtype T = T { getInt :: Int16 }
\end{code}
%
\noindent
Without shrinking the \ttp{Int16}s, the performance is dramatically improved.

Next, a programmer might try to control the complexity by truncating lists using
truncation where
%
\begin{code}
take n ls
\end{code}
%
returns the first $n$ elements \ttp{ls}.  The trade-off is quicker shrinking
with a lower probability of finding a smaller counterexample.  For example, we
might redefine shrink as follows:
%
\begin{samepage}
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- tk w
                , b <- tk i0, c <- tk i1
                , d <- tk i2, e <- tk i3 ]
  where tk x = take 10 (shrink x)
\end{code}
\end{samepage}
%
\noindent
Call this version ``truncated shrink''.  Truncation controls the blowup of the
input-space, but the downside is that potentially smaller counterexamples may be
omitted.

Someone who understands the semantics of QuickCheck's \ttp{shrink}
implementation defines a \ttp{shrink} instance as follows:\footnote{The
  following approach was suggested to the author by John Hughes.}
%
\begin{code}
shrink (T w i0 i1 i2 i3) = map go xs
  where xs = shrink (w, i0, i1, i2, i3)
        go (w', i0', i1', i2', i3')
           = T w' i0' i1' i2' i3'
\end{code}
%
\noindent
This ``tuple shrink'' definition does not suffer the same shortcomings: it
shrinks a value even if it contains an empty list, and the combinatorial blowup
of shrink candidates is avoided, since a pair \ttp{(a,b)} is shrunk by
attempting to shrink \ttp{a} while holding \ttp{b} constant, then attempting to
shrink \ttp{b} while holding \ttp{a} constant (and is generalized from pairs to
arbitrary tuples).  Thus, each argument to \ttp{T} is independently shrunk.
Still, using this definition, from some initial counterexamples, shrinking may
result in a final counterexample containing just over 100 \ttp{Int16} values.

@@

\begin{figure}[ht]
\includegraphics[scale=0.6]{../regression/PaperExample1/out/data}
%% \includegraphics[scale=0.6]{../regression/PaperExample1/out/time-big}
%% \includegraphics[scale=0.6]{../regression/PaperExample1/out/time-small}
  \caption{Results for 1000 tests of the program in Figure~\ref{fig:initial}.}
  \label{fig:graphs}
\end{figure}

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline
           & QC none     & QC trunc     & QC tuple    & SmartCheck\\
\hline \hline
Mean       & 81, 0.013s  & 40, 0.231s   & 13, 0.228s  & 6, 0.022s\\
\hline
Std. dev.  & 21, 0.001   & 21, 1.742    & 7,  0.962   & 4, 0.001s\\
% Median     & 0.009s, 67  & 0.17s, 33  & 1.81s, 13  & 0.10s, 6\\
%% \hline
%% MAD        & 0.02s, 13  & 0.09s, 8   & 0.62s, 3   & 0.02s, 2\\
\hline
95\%       & 116, 0.025s  & 84, 0.283s  & 26, 0.501s  & 13, 0.039s\\
\hline
    \end{tabular}
  \end{center}
  \caption{Summarizing data for the graph in Figure~\ref{fig:graphs}. Entries
    contain execution time (in seconds) and counterexample sizes (counting
    \ttp{Int16} values).}
  \label{table:results}
\end{table}

To motivate the benefits of SmartCheck reduction algorithm, consider
Table~\ref{table:results} and the corresponding graphs in
Figure~\ref{fig:graphs}, giving the results for using QuickCheck and SmartCheck
on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck, since as
noted, it is not even feasible for the example with default
instances).\footnote{All results reported in the paper are with a
  GHC-7.6.3-compiled program, using \ttp{-O2}, on a 2.8 GHz 4 core i7 with 16GB
  memory under OSX v. 10.9.2.}  We graph the final counterexample size and
execution time.

The experiments show three approaches to shrinking with QuickCheck: no shrinking
(QC none), to provide a baseline, truncated shrinking (QC trunc), as described
above, in which each list is truncated to 10 elements and \ttp{Int16} values are
not shrunk, and tuple shrinking (QC tuple), also described above, and
\ttp{Int16} values are also not shrunk.

With each shrinking strategy, we generate 1000 runs in which a random
counterexample is discovered (by QuickCheck), and then shrunk.  We plot the
number of runs resulting in a shrunk counterexample of a given size.

In Table~\ref{table:results}, we summarize the graphed values and provide
execution time in seconds as well.  For both execution time and final value
size, we summarize the mean, median, standard deviation, and the results at the
95th percentile.  (While we provide the standard deviations, note that the plots
are not necessarily Gaussian.)

The execution time includes the time to
discover an initial counterexample as well as time to shrink it.

SmartCheck produces smaller counterexamples than the other approaches, with a
very small tail.

@@

\multirow{4}{*}{Reverse}
  & Mean       & 2, 0.002s  & 2, \sci{4.8}{4}s  \\
  \cline{2-4}
  & Std. dev.  & 0, 0.002   & 0, \sci{8.1}{4}  \\
  \cline{2-4}
  & 95\%       & 2, 0.007s  & 2, \sci{5.8}{4}s  \\
  \hline

\multirow{4}{*}{Div0}
  & Mean       & 10, 0.001s  & 5, 0.001s  \\
  \cline{2-4}
  & Std. dev.  & 6, 0.002    & 0, 0.001   \\
  \cline{2-4}
  & 95\%       & 23, 0.002s  & 5, 0.002s \\
  \hline

\multirow{4}{*}{Heap}
  & Mean       & 15, 0.016s  & 7, 0.007s  \\
  \cline{2-4}
  & Std. dev.  & 1, 0.025    & 2, 0.003   \\
  \cline{2-4}
  & 95\%       & 19, 0.045s  & 12, 0.015s  \\
  \hline

\multirow{4}{*}{Parser}
  & Mean       & 191, 0.013s  & 16, 0.241s  \\
  \cline{2-4}
  & Std. dev.  & 195, 0.018   & 75, 0.800   \\
  \cline{2-4}
  & 95\%       & 577, 0.044s  & 14, 0.521s  \\
  \hline

    \end{tabular}

@@

\todo{note that 95 is smaller than average for heap}

@@

\section*{Acknowledgments}
I thank Andy Gill, Joe Hurd, John Launchbury, Simon Winwood, the anonymous
reviewers of ICFP'13 (from which an early draft of this paper was rejected), and
especially John Hughes for reading an earlier draft of this paper.  Their deep
insights and bug catches dramatically improved both the form and content.
